### 2022-10-04 AISTATS2023/paper.tex
explained better the variance part

\subsection{Interval-Based Formulation of ARE}
\label{sec:interval-based-estimation}

In real scenarios, the estimations are based on the limited instances
of the training set. Estimating \(\mu(z), \sigma(z)\) at the
granularity of a point is impossible, because the probability of
observing a sample inside the region \([x_s - h, x_s + h]\) tends to
zero, when \(h \to 0\). We define the region-effect and the
region-uncertainty inside the region \([z_1, z_2)\) as:

@@

\begin{equation}
  \label{eq:var_bin}
  \sigma(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 - z_1}}
\end{equation}

%
We also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
that partition the \(s\)-th feature axis into \(K\) variable-size
intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
\(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
redefine ARE using an interval-based formulation
\(\tilde{f}^{\mathtt{ARE}}(x_s):=
\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm
\tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

@@

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
\paragraph{Theorem 1.}
\label{sec:theorem-1}

\textit{If we define the residual \(\rho(z)\) as the difference
  between the expected effect at \(x_s\) and the region-effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the region-variance
  \(\sigma^2(z_1, z_2)\) equals to the integration of the point-wise
  variances plus the squared residuals inside the interval:
\begin{equation}
 \sigma^2(z_1, z_2) = \frac{\int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z}{z_2 - z_1}
\end{equation}}
%

\paragraph{Theorem 2.}
\label{sec:theorem-2}

\textit{If we define as \(\rho^2(z_{k-1}, z_{k})\) the
  region-residual, where
  \(\rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial
  z\), then the two definitions of ARE,
  \(f^{\mathtt{ARE}}, \tilde{f}^{\mathtt{ARE}}\) are equivalent, if
  and only if the sequence of limits \(\mathcal{Z}_s\) are such that
  \(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\).}

\paragraph{Discussion}

Theorem 1 decouples the aggregated uncertainty inside a bin, i.e.,
\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\), into two
terms. The first term quantifies the accumulated uncertainty due to
the natural characteristics of the experiment, i.e.,
\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}^2(z) \partial z\),
and the second term adds extra nuisance uncertainty due to limiting
the resolution
\(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
\). Therefore, it holds that:
\begin{equation}
  \label{eq:uncertainty_bin}
  \mathcal{H}_{bin}(z_1, z_2) = \mathcal{H}(z_1, z_2) + \mathcal{H}_n(z_1, z_2)
\end{equation}
%
In Theorem 2 we prove that if we find a sequence of bins, where each
bin adds no nuisance uncertainty, then we will get exactly the same
ARE explanation as in the original definition. In general, we use the
quantity \(Q_{\mathcal{Z}} = \sum_{k=1}^K \rho^2(z_{k-1}, z_{k})\) to
measure the dissimilarity between the actual definition of ARE
\(f^{\mathtt{ARE}}\) and the region-based definition
\(\tilde{f}^{\mathtt{ARE}}\) using the sequence of intervals
\(\mathcal{Z}_s\). In this way, we access a quantified metric for
evaluating how good is a particular splitting.

@@

In this section we formulate bin-splitting as an unsupervised
clustering problem. The mean effect (Eq.~\eqref{eq:mu_bin}) and the
uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
the set \(\mathcal{S}_k\) of the dataset instances with the \(s\)-th
feature lying inside the interval
\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \} \):

% Under this
% assumption, and due to the law of large numbers, the approximations
% become more robust, as the bins become wider. On the other hand, bins
% should group local explanations with similar effects and enlarging a
% bin is under the danger of adding nuisance uncertainty.

Both approximations are unbiased only under the assumption that the
points are uniformly distributed inside the intervals. We set-up the
following optimization problem:

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - 0.2 \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of bin-costs. The cost of each bin is the mean
variance \(\hat{sigma}^2_k\) scaled by the bin length \(\Delta z_k\)
and discounted by the term \(\tau_k\). The term \(\tau_K\) acts as a
tie-braker in cases of two candidate bins with equal mean variance,
favoring the selection of the bigger bin. The constraint of at least
\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
\textit{robust} estimation.

According to Eq.~(\ref{eq:uncertainty_bin}), the uncertainty of a bin
\(\mathcal{H}_{bin}\) is the sum of the uncertainty due to the
characteristics of the experiment (data and black-box model)
\(\mathcal{H}\) and the nuisance uncertainty \(\mathcal{H}_n\). Since
\(\mathcal{H}\) is independent of the bin-spliting procedure, by
minimizing \(\mathcal{L}\) we minimize the nuiscane uncertainty
\(\mathcal{H}_n\) in each bin.

>>>>

\subsection{Interval-Based Formulation and Approximation}
\label{sec:interval-based-estimation}

In real scenarios, the estimations are based on the limited instances
of the training set. Estimating \(\mu(z), \sigma(z)\) at the
granularity of a point is impossible, because the probability of
observing a sample inside the region \([x_s - h, x_s + h]\) tends to
zero, when \(h \to 0\). Therefore we define the regional-effect
\(\mu(z_1, z_2)\) and the regional-uncertainty \(\sigma(z_1, z_2)\),
as the summary statistics that characterize the effect and the
uncertainty inside the interval \([z_1, z_2)\):

@@

\begin{equation}
  \label{eq:var_bin}
  \sigma(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z) )^2 \right] \partial z}{z_2 - z_1}}
\end{equation}

%
The regional-effect and the regional-uncertainty are the mean values
of the local effects and local uncertainty inside an interval. In
other words, they quantify what is the expected effect and the
expected uncertainty of a sample drawn at random from \([z_1, z_2]\)
given uniform distribution \(\mathcal{U}(z_1, z_2)\).

If we also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
that partition the \(s\)-th feature axis into \(K\) variable-size
intervals, i.e. \(\mathcal{Z}_s = \{z_0, \ldots, z_K\}\), we can
redefine ARE using an interval-based formulation
\(\tilde{f}^{\mathtt{ARE}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)
\pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

@@

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%

\subsection{Interval-Based Approximation}

The mean effect (Eq.~\eqref{eq:mu_bin}) and the
uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
the set \(\mathcal{S}_k\) of dataset instances with the \(s\)-th
feature lying inside the interval
\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \} \):

Under the assumption that the points are uniformly distributed inside
the intervals Eq.\eqref{eq:2} is an unbiased estimatior of
Eq.~(\ref{eq:mu_bin}). The same does not hold in the general case for
Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
is an unbiased estimator of the observable standard deviation
\(\sigma_{obs}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
    \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2
    \right] \partial z}{z_2 - z_1}}\), which as we prove in Theorem 1
is an overestimation of the real regional variance.

\paragraph{Theorem 1.}
\label{sec:theorem-1}

\textit{If we define the residual \(\rho(z)\) as the difference
  between the expected effect at \(x_s\) and the regional effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the regional variance
  \(\sigma^2(z_1, z_2)\) equals to:}

\begin{equation}
    \label{eq:bin-uncertainty-proof}
 \sigma_{obs}^2(z_1, z_2) = \sigma^2(z) + \frac{\int_{z_1}^{z_2}\rho^2(z) \partial z}{z_2 - z_1}
\end{equation}
\noindent
Theorem 1 reveals an important connection between the ground-truth and
the observable uncertainty inside a region. Setting as ground-truth
the regional uncertainty
\(\mathcal{H}_{true}(z_1, z_2) = \sigma(z_1, z_2)}\) and as observable
uncertainty the one we estimate with Eq.~(\ref{eq:3}), i.e.
\(\mathcal{H}_{obs}(z_1, z_2) := \sigma_{obs}(z_1, z_2)\), then there
is an error term quantified by the residual inside the interval, i.e.
\(\mathcal{E}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}\rho^2(z)
    \partial z}{z_2 - z_1}\)}, such that:

\begin{equation}
  \label{eq:uncertainty-bin}
  \mathcal{H}^2_{obs}(z_1, z_2) = \mathcal{H}^2_{true}(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
\end{equation}
%
Eq.~(\ref{eq:uncertainty-bin}) shows that the error term quantifies
how much the observable uncertainty of the region \([z_1, z_2]\)
deviates for the one we would like to measure and provides a
quantified metric to evaluate a particular splitting. As we prove in
the Appendix, UALE uncertainty approximation is unbiased if and only if the
partitioning is such that ther error term is zero for all bins, i.e. 
\(\sum_{k=1}^K \mathcal{E}^2(z_{k-1}, z_{k}) = 0\)

@@

In this section we formulate bin-splitting as an unsupervised
clustering problem. The main idea of the clustering problem is to find
the best bin-splitting solution, handling appropriately two
conflicting objectives. On the one hand, we want to minimize the error
term \(\mathcal{E}\) defined above for minimizing the error in the
uncertainty estimation and on the other hand, we want wide-enough bins
for including a fair population of samples for a robust estimation of
\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\). We set-up the
following optimization problem:

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of bin costs. The cost of each bin is the
approximated variance \(\hat{\sigma}^2_k\) scaled by the bin length
\(\Delta z_k\) and discounted by the term \(\tau_k\). The term
\(\tau_K\) acts as a tie-braker that favors the selection of a bigger
bin in case it has similar variance with the aggregate variance of
many smaller ones. The constraint of at least \(N_{\mathtt{PPB}}\)
points per bin sets the lowest-limit for a \textit{robust}
estimation. The user can choose to what extent they favor the creation
of wide bins through the discount parameter by setting the parameter
\(\alpha\) and where they set the minimum for robust approximation
with the parameter \(N_{\mathtt{PPB}}\). For providing a rough idea of
some sensible value, in our experiments we set \(\alpha = 0.2\) which
means that the discount ranges in \([0\%, 20\%]\) and
\(N_{\mathtt{PPB}} = \frac{N}{20}\). It is also important to clarify
that by minimizing \(\hat{\sigma}^2_k \approx \mathcal{H}_{obs}^2\),
we actually minimize the squared error term \(\mathcal{E}^2\), since
the term \(\mathcal{H}^2_{true}\) is independent of the bin splitting
procedure.
